---
title: "ABCD_Participants_Imaging_Demographics_OK"
author:
  - name: Hang Yang
    orcid: 0000-0002-7349-3366
    email: yanghang@cibr.ac.cn
    url: https://scholar.google.com/citations?user=LqI72qAAAAAJ&hl
    affiliation: 
      - name: Chinese Institute for Brain Research, Beijing (CIBR)
        group: CUIZAIXU Lab
        url: https://cuilab.cibr.ac.cn/
date: "2026-01-05"
format: 
  html:
    code-fold: true
    code-tools: true
    theme: Materia
    toc: true
    toc_depth: 3
    toc-expand: 1
    toc_float: true
    toc-location: left
    number-sections: true
    number-depth: 3
    embed-resources: true
---

# Diagnosis

In this work, we focused on the attention-deficit/hyperactivity disorder (**adhd**) from the Adolescent Brain Cognitive Development (ABCD) Study. Specifically, the ADHD participants were mainly determined based on ([Norman et al., 2024](https://psychiatryonline.org/doi/10.1176/appi.ajp.20230026)):

Met ADHD-present, partial remission or unspecified on the KSADS-COMP. Exclude ADHD-past-only.

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
})

rm(list = ls())

demo_path <- "D:/ABCD/abcd-data-release-5.1/core/"
behav_path <- "F:/Cui_Lab/Projects/ABCD_ADHD_Control_Energy/data/sub_info/"

# Helper Functions ----
# Track exclusions at each step
track_exclusions <- function(data, prev_counts, step_name) {
  curr_0y <- sum(data$eventname == "baseline_year_1_arm_1")
  curr_2y <- sum(data$eventname == "2_year_follow_up_y_arm_1")
  
  cat(sprintf("%s (baseline): exclude %d participants\n", 
              step_name, prev_counts$n_0y - curr_0y))
  cat(sprintf("%s (2-year follow-up): exclude %d participants\n", 
              step_name, prev_counts$n_2y - curr_2y))
  
  list(n_0y = curr_0y, n_2y = curr_2y)
}

# Get current sample counts
get_counts <- function(data) {
  list(
    n_0y = sum(data$eventname == "baseline_year_1_arm_1"),
    n_2y = sum(data$eventname == "2_year_follow_up_y_arm_1")
  )
}
```

## Imporat KASDS data

```{r ksads}
mh_p_ksads <- fread(paste0(demo_path, "mental-health/mh_p_ksads_ss.csv"))
mh_y_ksads <- fread(paste0(demo_path, "mental-health/mh_y_ksads_ss.csv"))

mh_p_ksads <- mh_p_ksads %>% filter(eventname %in% c("baseline_year_1_arm_1", "2_year_follow_up_y_arm_1"))
mh_y_ksads <- mh_y_ksads %>% filter(eventname %in% c("baseline_year_1_arm_1", "2_year_follow_up_y_arm_1"))
```

## ADHD diagnosis

ADHD: present (ksads_14_853_p) or partial remission (ksads_14_855_p) or unspecified (ksads_14_856_p).

```{r adhd}
# Parent
adhd_mh_p_ksads <- mh_p_ksads %>%
  filter(ksads_14_853_p == 1 | ksads_14_855_p == 1 | ksads_14_856_p == 1) %>%
  select(src_subject_id, eventname, ksads_14_853_p, ksads_14_855_p, ksads_14_856_p)

# Get ADHD subjects at baseline (Year 0)
sub_adhd_0y <- adhd_mh_p_ksads %>%
  filter(eventname == "baseline_year_1_arm_1") %>% 
  select(src_subject_id, eventname)

# Load longitudinal tracking data to identify follow-up visits
abcd_y_lt <- fread(paste0(demo_path, "abcd-general/abcd_y_lt.csv"))

# Get 2-year follow-up data ONLY for subjects who were ADHD at baseline
sub_adhd_2y <- abcd_y_lt %>% 
  filter(src_subject_id %in% sub_adhd_0y$src_subject_id & eventname == "2_year_follow_up_y_arm_1") %>% 
  select(src_subject_id, eventname)

# Combine baseline and follow-up ADHD subjects
sub_adhd <- bind_rows(sub_adhd_0y, sub_adhd_2y)
sub_adhd$diagnosis <- "adhd"
```

## Typically developing children

Typically developing (TD) children, without any diagnosis.

```{r td}
# Define function to select TD subjects for a given event and return with eventname
select_td <- function(mh_p_data, mh_y_data, mh_p_dict, mh_y_dict, event_name) {
  
  # Filter for the given event
  mh_p <- filter(mh_p_data, eventname == event_name)
  mh_y <- filter(mh_y_data, eventname == event_name)
  
  # Get unique subjects for this event
  sub_all <- unique(c(mh_p$src_subject_id, mh_y$src_subject_id))
  
  # Select diagnosis columns
  td_p_ksads <- mh_p %>% select(one_of(mh_p_dict$var_name))
  td_y_ksads <- mh_y %>% select(one_of(mh_y_dict$var_name))
  
  # Replace NAs with 0 and check for diagnoses
  td_p_ksads[is.na(td_p_ksads)] <- 0
  td_y_ksads[is.na(td_y_ksads)] <- 0
  td_p_ksads <- as.data.frame(td_p_ksads == 1)
  td_y_ksads <- as.data.frame(td_y_ksads == 1)
  
  # Identify subjects with diagnoses
  sub_ksads_p <- mh_p$src_subject_id[rowSums(td_p_ksads) > 0]
  sub_ksads_y <- mh_y$src_subject_id[rowSums(td_y_ksads) > 0]
  
  # Exclude all diagnosed subjects in ksads for this event
  sub_ksads <- unique(c(sub_ksads_p, sub_ksads_y))
  sub_td <- setdiff(sub_all, sub_ksads)
  
  # Combine TD subjects with eventname
  td_subjects <- data.frame(src_subject_id = sub_td, eventname = event_name)
  
  return(td_subjects)
}

# Load the diagnosis dictionaries
mh_p_ksads_dict <- fread(paste0(behav_path, "diagnosis/mh_p_ksads_ss-ABCD-Data-Dictionary.csv")) %>%
  filter(table_name_nda == "abcd_ksad01", grepl("Diagnosis - ", var_label))
mh_y_ksads_dict <- fread(paste0(behav_path, "diagnosis/mh_y_ksads_ss-ABCD-Data-Dictionary.csv")) %>%
  filter(table_name_nda == "abcd_ksad501", grepl("Diagnosis - ", var_label))

# Select TD subjects for baseline and follow-up, with eventname
sub_td_0y <- select_td(mh_p_ksads, mh_y_ksads, mh_p_ksads_dict, mh_y_ksads_dict, "baseline_year_1_arm_1")
sub_td_2y <- select_td(mh_p_ksads, mh_y_ksads, mh_p_ksads_dict, mh_y_ksads_dict, "2_year_follow_up_y_arm_1")

# Exclude any subjects who are in the ADHD group
sub_td_0y <- sub_td_0y %>% 
  filter(! src_subject_id %in% sub_adhd$src_subject_id)
sub_td_2y <- sub_td_2y %>% 
  filter(! src_subject_id %in% sub_adhd$src_subject_id)

# Combine baseline and follow-up TD subjects
sub_td <- bind_rows(sub_td_0y, sub_td_2y)
sub_td$diagnosis <- "td"
```

## Combine TD and ADHD

```{r combine_td_adhd}
sub_info <- bind_rows(sub_td, sub_adhd)
sub_info <- merge(sub_info, mh_p_ksads[, c("src_subject_id", "eventname", "ksads_14_853_p", "ksads_14_855_p", "ksads_14_856_p")], 
                  by = c("src_subject_id", "eventname"), all.x = TRUE)

sub_info <- sub_info %>% 
  rename(
    ksads_adhd_present = ksads_14_853_p,
    ksads_adhd_remission = ksads_14_855_p,
    ksads_adhd_unspecified = ksads_14_856_p
  )

sub_info <- sub_info %>%
  mutate(
    ksads_adhd = case_when(
      ksads_adhd_present == 1 ~ "present",
      ksads_adhd_remission == 1 ~ "remission",
      ksads_adhd_unspecified == 1 ~ "unspecified",
      TRUE ~ NA_character_ # Handle cases where none of the conditions are met
    )
  )

sub_info$eventname <- factor(sub_info$eventname, 
                             levels = c("baseline_year_1_arm_1", "2_year_follow_up_y_arm_1"))

N1 <- dim(sub_info)[1]
N1_0y <- dim(filter(sub_info, sub_info$eventname == "baseline_year_1_arm_1"))[1]
N1_2y <- dim(filter(sub_info, sub_info$eventname == "2_year_follow_up_y_arm_1"))[1]
```

::: {.callout-note icon="false"}
`r N1` visits enrolled.

`r knitr::kable(table(sub_info[, c("eventname", "diagnosis")]), align = 'l') %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left")`
:::

# Imaging

## QC

Include only scans that passed QC for BOTH:

-   T1-weighted structural imaging (imgincl_t1w_include == 1)
-   Diffusion MRI (imgincl_dmri_include == 1)

```{r QC}
counts <- get_counts(sub_info)

mri_y_qc_incl <- fread(paste0(demo_path, "imaging/mri_y_qc_incl.csv"))

sub_qc_incl <- mri_y_qc_incl %>% 
  filter(eventname %in% c("baseline_year_1_arm_1", "2_year_follow_up_y_arm_1")) %>%
  filter(imgincl_t1w_include == 1 & imgincl_dmri_include == 1) %>% 
  select(src_subject_id, eventname)

sub_info <- merge(sub_info, sub_qc_incl, by = c("src_subject_id", "eventname"))
counts <- track_exclusions(sub_info, counts, "Not meet official quality control")
```

## Preprocessing

-   Finished DWI data preprocessing and structural connectome reconstruction.
-   No isolated brain regions in SC network.

```{r qsiprep}
counts <- get_counts(sub_info)

# load subject list after finish qsiprep preprocessing
sub_dwi_0y <- fread(paste0(behav_path, "sub_dwi_strength_fd_baselineYear1Arm1_qc.csv"))
sub_dwi_0y$src_subject_id <- sub_dwi_0y$sub_id  %>% 
  gsub("_ses-baselineYear1Arm1", "", .) %>% 
  gsub("sub-NDAR", "NDAR_", .)
sub_dwi_0y$eventname <- "baseline_year_1_arm_1"

sub_dwi_2y <- fread(paste0(behav_path, "sub_dwi_strength_fd_2YearFollowUpYArm1_qc.csv"))
sub_dwi_2y$src_subject_id <- sub_dwi_2y$sub_id  %>% 
  gsub("_ses-2YearFollowUpYArm1", "", .) %>% 
  gsub("sub-NDAR", "NDAR_", .)
sub_dwi_2y$eventname <- "2_year_follow_up_y_arm_1"

sub_dwi <- bind_rows(sub_dwi_0y, sub_dwi_2y)

# interaction
sub_info <- merge(sub_info, sub_dwi, by = c("src_subject_id", "eventname"))
counts <- track_exclusions(sub_info, counts, "Failed preprocessing and reconstruction, with isolated regions")
```

## HeadMotion

Head motion: Less than mean + 3\*SD

```{r headmotion}
counts <- get_counts(sub_info)

## exclude FD > mean + 3*SD
mean_fd <- sub_info$mean_fd
fd_thr <- mean(mean_fd) + 3*sd(mean_fd)

sub_info <- filter(sub_info, mean_fd < fd_thr)
counts <- track_exclusions(sub_info, counts, "Excessive head motion: FD > mean + 3×SD")

N2 <- dim(sub_info)[1]
N2_0y <- dim(filter(sub_info, sub_info$eventname == "baseline_year_1_arm_1"))[1]
N2_2y <- dim(filter(sub_info, sub_info$eventname == "2_year_follow_up_y_arm_1"))[1]
```

::: {.callout-note icon="false"}
`r N2` scans successfully completed SC network reconstruction. A total of `r N1 - N2` scans were excluded (`r N1_0y - N2_0y` at baseline, `r N1_2y - N2_2y` at 2-year follow-up).

`r knitr::kable(table(sub_info[, c("eventname", "diagnosis")]), align = 'l') %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left")`
:::

# Demographics

## Age

Exclude participants with invalid Age (NA).

```{r Age}
counts <- get_counts(sub_info)

abcd_y_lt <- fread(paste0(demo_path, "abcd-general/abcd_y_lt.csv"))
# Month to Year
abcd_y_lt$interview_age <- abcd_y_lt$interview_age/12

sub_info <- merge(sub_info, abcd_y_lt[, c("src_subject_id", "eventname", "interview_age")], 
                  by = c("src_subject_id", "eventname"))
sub_info <- filter(sub_info, !is.na(interview_age))
sub_info <- rename(sub_info, age = interview_age)

counts <- track_exclusions(sub_info, counts, "Invalid age")
```

## Sex

Exclude participants with invalid Sex (3, 4, 999).

1 = Male; 2 = Female; 3 = Intersex-Male; 4 = Intersex-Female; 999 = Don"t know.

```{r Sex}
counts <- get_counts(sub_info)

abcd_p_demo <- fread(paste0(demo_path, "abcd-general/abcd_p_demo.csv"))
sub_info <- sub_info %>% left_join(select(abcd_p_demo, c("src_subject_id", "demo_sex_v2")), 
                                   by = "src_subject_id", multiple="first")
sub_info <- filter(sub_info, demo_sex_v2 == 1 | demo_sex_v2 == 2)

# Convert from numbers to characters
sub_info <- sub_info %>%
  mutate(sex = case_when(
    demo_sex_v2 == 1 ~ "Male", 
    demo_sex_v2 == 2 ~ "Female", 
    TRUE ~ NA_character_
  ))

sub_info <- sub_info %>% select(-demo_sex_v2)
counts <- track_exclusions(sub_info, counts, "Invalid sex")
```

## Race/Ethnicity

Exclude participants with invalid Race/Ethnicity (NA).

1 = White; 2 = Black; 3 = Hispanic; 4 = Asian; 5 = Other.

```{r Race}
counts <- get_counts(sub_info)

abcd_p_demo <- fread(paste0(demo_path, "abcd-general/abcd_p_demo.csv"))
sub_info <- sub_info %>% left_join(select(abcd_p_demo, c("src_subject_id", "race_ethnicity")), 
                                   by = "src_subject_id", multiple = "first")

sub_info <- sub_info %>% 
  filter(!is.na(race_ethnicity))

# Convert from numbers to characters
sub_info <- sub_info %>%
  mutate(race = case_when(
    race_ethnicity == 1 ~ "White", 
    race_ethnicity == 2 ~ "Black", 
    race_ethnicity == 3 ~ "Hispanic", 
    race_ethnicity == 4 ~ "Asian", 
    race_ethnicity == 5 ~ "Other", 
    TRUE ~ NA_character_
  ))

sub_info <- sub_info %>% select(-race_ethnicity)
counts <- track_exclusions(sub_info, counts, "Invalid race/ethnicity")
```

## Handedness

Exclude participants with invalid Handedness (NA).

1=right handed; 2=left handed; 3=mixed handed.

```{r Handedness}
counts <- get_counts(sub_info)

nc_y_ehis <- fread(paste0(demo_path, "neurocognition/nc_y_ehis.csv"))
sub_info <- sub_info %>% left_join(select(nc_y_ehis, c("src_subject_id", "ehi_y_ss_scoreb")), 
                                   by = "src_subject_id", multiple = "first")
sub_info <-  sub_info %>% 
  filter(!is.na(ehi_y_ss_scoreb))

# Convert from numbers to characters
sub_info <- sub_info %>%
  mutate(handedness = case_when(
    ehi_y_ss_scoreb == 1 ~ "right", 
    ehi_y_ss_scoreb == 2 ~ "left", 
    ehi_y_ss_scoreb == 3 ~ "mixed", 
    TRUE ~ NA_character_
  ))

sub_info <- sub_info %>% select(-ehi_y_ss_scoreb)
counts <- track_exclusions(sub_info, counts, "Invalid handedness")
```

## Site/Scanner

Exclude participants with invalid Site (NA).

```{r site}
counts <- get_counts(sub_info)

# site
sub_info <- merge(sub_info, abcd_y_lt[, c("src_subject_id", "eventname", "site_id_l")], 
                  by = c("src_subject_id", "eventname"))

sub_info <- sub_info %>% rename(site_id = site_id_l)
sub_info$site_id2 <- gsub("site0", "site", sub_info$site_id)

# Inclusion
sub_info <- filter(sub_info, !is.na(site_id))
counts <- track_exclusions(sub_info, counts, "Invalid site/scanner")
```

## Count

For the model to accurately parse variance attributable to age, sex and site, sufficient observations are required for any given combination of these variables. We therefore excluded data cells containing less than ten HCs for a particular sex at a given site. If this exclusion procedure resulted in less than ten cases in total from any given site, the entire site was excluded ([Segal et al., 2023](https://www.nature.com/articles/s41593-023-01404-6)):

```{r count}
counts <- get_counts(sub_info)

table(filter(sub_info, diagnosis == "td") %>% select(site_id, sex))
# Inclusion
sub_info <- filter(sub_info, site_id != "site22")
counts <- track_exclusions(sub_info, counts, "Sites with less than ten TDCs for a particular sex")
```

## TBV

Total brain volume (TBV) is included as a covariate.

```{r tbv}
counts <- get_counts(sub_info)

# TBV
mri_y_smr_vol_aseg <- fread(paste0(demo_path, "imaging/mri_y_smr_vol_aseg.csv"))

sub_info <- merge(sub_info, mri_y_smr_vol_aseg[, c("src_subject_id", "eventname", "smri_vol_scs_wholeb")], 
                  by = c("src_subject_id", "eventname"))

sub_info <- sub_info %>% 
  rename(
    network_strength_schaefer400 = schaefer400,
    network_strength_schaefer200 = schaefer200,
    TBV = smri_vol_scs_wholeb
  )

sub_info$TBV <- sub_info$TBV / 1000 # mm^3 to mL

sub_info <- filter(sub_info, !is.na(TBV))
counts <- track_exclusions(sub_info, counts, "Invalid TBV")
```

# Psychopathology

## Child Behavior Checklist (CBCL)

The **Child Behavior Checklist** (**CBCL**) is a widely used caregiver report form identifying problem behavior in children. Here, we used the **t scores**.

```{r cbcl}
mh_p_cbcl <- fread(paste0(demo_path, "mental-health/mh_p_cbcl.csv"))
mh_p_cbcl <- as.data.frame(mh_p_cbcl)

mh_p_cbcl <- mh_p_cbcl %>%
  select(src_subject_id, eventname, matches("^cbcl_scr_syn_.*_t$"))
# mh_p_cbcl <- mh_p_cbcl %>% 
#   select(c("src_subject_id", "eventname", ends_with("_t")))

CBCL_items <- mh_p_cbcl %>% 
  select(ends_with("_t")) %>% 
  select(-starts_with("cbcl_scr_07")) %>% 
  colnames()

sub_info <- merge(sub_info, mh_p_cbcl[, c("src_subject_id", "eventname", CBCL_items)], 
                  by = c("src_subject_id", "eventname"), all.x = TRUE)
```

# Save

Filter dataset to include:

-   All TD subjects (any timepoint)
-   ADHD subjects at baseline
-   ADHD subjects at 2-year follow-up ONLY if they also have baseline data

```{r save_behav}
# Exclude ADHD only have 2-year data
adhd_baseline_ids <- sub_info %>%
  filter(diagnosis == "adhd", eventname == "baseline_year_1_arm_1") %>%
  pull(src_subject_id) %>%
  unique()

sub_info <- sub_info %>%
  filter(
    diagnosis == "td" |
    (diagnosis == "adhd" & eventname == "baseline_year_1_arm_1") |
    (diagnosis == "adhd" & eventname == "2_year_follow_up_y_arm_1" & src_subject_id %in% adhd_baseline_ids)
  )

# sort subjects based on src_subject_id and eventname
sub_info <- sub_info %>%
  arrange(src_subject_id, factor(eventname, levels = c("baseline_year_1_arm_1", "2_year_follow_up_y_arm_1")))

sub_info <- sub_info %>%
  mutate(eventname2 = case_when(
    eventname == "baseline_year_1_arm_1" ~ "baselineYear1Arm1", 
    eventname == "2_year_follow_up_y_arm_1" ~ "2YearFollowUpYArm1", 
    TRUE ~ NA_character_
  ))

sub_info %>%
  filter(diagnosis %in% c("td", "adhd")) %>% 
  group_by(diagnosis) %>%
  summarise(
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  )

fwrite(sub_info, paste0(behav_path, "sub_demo_info.csv"), row.names = FALSE)

N3 <- dim(sub_info)[1]
N3_0y <- dim(filter(sub_info, sub_info$eventname == "baseline_year_1_arm_1"))[1]
N3_2y <- dim(filter(sub_info, sub_info$eventname == "2_year_follow_up_y_arm_1"))[1]
```

::: {.callout-note icon="false"}
`r N3` scans include. A total of `r N2 - N3` scans were excluded (`r N2_0y - N3_0y` at baseline, `r N2_2y - N3_2y` at 2-year follow-up).

`r knitr::kable(table(sub_info[, c("eventname", "diagnosis")]), align = 'l') %>% kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left")`
:::
